Wine Quality by Alona Varshal
========================================================

Having worked previously on wines, I chose the red and white wines data set. I combined the two using the rbind() command after removing the variable "X", adding a variable "type" having a value of 1 for red wines and 0 for white wines. The new data set, wines, have 13 features which include type and quality which were both converted from integer to factor variables. The goal of this exploratory analysis is to determine which features contribute to the most separation of the different values of quality of wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(gridExtra)
library(psych)
library(plyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd('~/DataScience/udacity/dand/explore_summarize_data/project/')
redwine <- read.csv('wineQualityReds.csv')
whitewine <- read.csv('wineQualityWhites.csv')
```

```{r echo=FALSE, removing_X}
redwine$X <- NULL
whitewine$X <- NULL
```

```{r echo=FALSE, adding_type}
redwine$type <- 1
whitewine$type <- 0
```

```{r echo=FALSE, combining_wines}
wines <- rbind(redwine, whitewine)
```

```{r echo=FALSE, convert_type_factor}
wines$type <- factor(wines$type)
```

```{r echo=FALSE, Change_quality_to_Factor_Variable}
wines$quality <- factor(wines$quality)
```

### Structure of wines data frame

```{r echo=FALSE, Structure_data_frame}
str(wines)
```
### Summary of data frame
```{r echo=FALSE, Summary}
summary(wines)
```
### Description of data frame
```{r}
describe(wines)
```

# Univariate Plots Section

### Distribution of wines according to quality
```{r echo=FALSE, histograms}
ggplot(aes(x = quality), data = wines) +
  geom_histogram(stat = 'count', fill = I('blue')) +
  ggtitle('Distribution of Wine Qualities') + 
  theme(text = element_text(size = 10))

ggplot(aes(x = quality), data = wines) +
  geom_histogram(stat = 'count', aes(fill = type)) +
  scale_fill_brewer(type = 'qual') +
  facet_wrap( ~ type) +
  ggtitle('Distribution of Qualities of White (0) and Red (1) Wines in the Wine Quality Data Set') +
  theme(text = element_text(size = 10))
```

### Distribution of wine features
```{r echo=FALSE, wines_features_histograms}
fixedacidityp1 <- ggplot(aes(x = fixed.acidity), data = wines) +
  geom_histogram(fill = I('#ddb40d'), color = I('black'), binwidth = 0.5) +
  scale_x_continuous(breaks = seq(4, 18, 2)) +
  ggtitle('Fixed Acidity') +
  theme(text = element_text(size=10)) 

volatileacidityp1 <- ggplot(aes(x = volatile.acidity), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#d9e570'), binwidth = 0.05) +
  scale_x_continuous(limits = c(0, 1)) +
  ggtitle('Volatile Acidity') +
  theme(text = element_text(size=10))

citricacidp1 <- ggplot(aes(x = citric.acid), data = wines) +
  geom_histogram(color = I('black'), fill = I ('orange')) +
  scale_x_continuous(limits = c(0, 1)) +
  ggtitle('Citric Acid') +
  theme(text = element_text(size=10))

residualsugarp1 <- ggplot(aes(x = residual.sugar), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#35a1ba'), binwidth = 0.2) +
  scale_x_continuous(breaks = seq(0, 7, 1), limits = c(0, 7)) +
  ggtitle('Residual Sugar') +
  theme(text = element_text(size=10))

chloridesp1 <- ggplot(aes(x = chlorides), data = wines) +
  geom_histogram(color = I('black'), fill = I ('yellow'), binwidth = 0.005) +
  scale_x_continuous(limits = c(0.025, 0.15), breaks = seq(0.03, 0.15, 0.03)) +
  ggtitle('Chlorides') +
  theme(text = element_text(size=10))

freeSO2p1 <- ggplot(aes(x = free.sulfur.dioxide), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#35ba5d')) +
  ggtitle('Free Sulfur Dioxide') +
  scale_x_continuous(breaks = seq(0, 150, 50), limits = c(0, 150)) +
  theme(text = element_text(size=10))

totalSO2p1 <- ggplot(aes(x = total.sulfur.dioxide), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#5d6d3a'), binwidth = 10) +
  scale_x_continuous(limits = c(0, 300)) +
  ggtitle('Total Sulfur Dioxide') +
  theme(text = element_text(size=10))

densityp1 <- ggplot(aes(x = density), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#1f61c4'), binwidth = 0.001) +
  scale_x_continuous(limits = c(0.985, 1.005)) +
  ggtitle('Density') +
  theme(text = element_text(size=10))

pHp1 <- ggplot(aes(x = pH), data = wines) +
  geom_histogram(color = I('black'), fill = I ('gray'), binwidth = 0.05) +
  ggtitle('pH') +
  theme(text = element_text(size=10))

SO4p1 <- ggplot(aes(x = sulphates), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#1fc42f'), binwidth = 0.05) +
  ggtitle('Sulphates') +
  scale_x_continuous(limits = c(0.25,1.5)) +
  theme(text = element_text(size=10))

alcoholp1 <- ggplot(aes(x = alcohol), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#727bd8'), binwidth = 0.2) +
  ggtitle('Alcohol') +
  theme(text = element_text(size=10))

grid.arrange(fixedacidityp1, volatileacidityp1, citricacidp1, residualsugarp1, chloridesp1, freeSO2p1, ncol=3)

grid.arrange(totalSO2p1, densityp1, pHp1, SO4p1, alcoholp1, ncol = 3)
```

### Transformations

All the non-normal distributions were transformed in some way.

Fixed acidity is tailing, so transformation was done. Resulting plot looks a little better, but not too much.
```{r echo=FALSE, transformations_histograms}
fixedacidityp2 <- ggplot(aes(x = fixed.acidity), data = wines) +
  geom_histogram(fill = I('#ddb40d'), color = I('black')) +
  scale_x_log10() +
  ggtitle('Fixed Acidity, log10') +
  theme(text = element_text(size=10)) 

grid.arrange(fixedacidityp1, fixedacidityp2, ncol = 2)
```

Volatile acidity is bimodal. Transformation didn't do much to change it.
```{r echo=FALSE}
volatileacidityp2 <- ggplot(aes(x = volatile.acidity), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#d9e570'), binwidth = 0.05) +
  scale_x_log10(breaks = c(0.1, 0.2, 0.5, 1.0)) +
  ggtitle('Volatile Acidity, log10') +
  theme(text = element_text(size=10))

grid.arrange(volatileacidityp1, volatileacidityp2, ncol = 2)

# citric acid is good, only has outliers in the upper range. Same as fixed acidity.
```


Residual sugar is not normally distributed. Transformation using log10 yields something like a bimodal distribution.
```{r echo=FALSE}
residualsugarp2 <- ggplot(aes(x = residual.sugar), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#35a1ba'), binwidth = 0.05) +
  scale_x_log10(breaks = c(1,2, 5, 10, 25, 50, 75)) +
  ggtitle('Residual Sugar, log10') +
  theme(text = element_text(size=10))

grid.arrange(residualsugarp1, residualsugarp2, ncol = 2)
```

Chlorides also don't look normally distributed. Transformation made it look better.
```{r echo=FALSE}
chloridesp2 <- ggplot(aes(x = chlorides), data = wines) +
  geom_histogram(color = I('black'), fill = I ('yellow')) +
  scale_x_log10(breaks = c(0.01, 0.02, 0.05, 0.10, 0.2, 0.5)) +
  ggtitle('Chlorides, log10') +
  theme(text = element_text(size=10))

grid.arrange(chloridesp1, chloridesp2, ncol = 2)

# Free sulfur dioxide does not look like a normal curve but transformation didn't result into a normal distribution.
freeSO2p2 <- ggplot(aes(x = free.sulfur.dioxide), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#35ba5d')) +
  scale_x_log10() +
  ggtitle('Free Sulfur Dioxide') +
  theme(text = element_text(size=10))

# Total sulfur dioxide distribution is bimodal and transformation didn't result into better curves.
totalSO2p2 <- ggplot(aes(x = total.sulfur.dioxide), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#5d6d3a')) +
  scale_x_sqrt() +
  ggtitle('Total Sulfur Dioxide') +
  theme(text = element_text(size=10))

# Transformation of density also didn't make the distribution better.
densityp2 <- ggplot(aes(x = density), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#1f61c4'), binwidth = 0.0005) +
  scale_x_log10() +
  ggtitle('Density, log10') +
  theme(text = element_text(size=10))

# Alcohol distribution isn't normal and transformation didn't do anything to improve the plot.
alcoholp2 <- ggplot(aes(x = alcohol), data = redwine) +
  geom_histogram(color = I('black'), fill = I ('#727bd8')) +
  scale_x_log10() +
  ggtitle('Alcohol, log10') +
  theme(text = element_text(size=10))
```

Calculating the ratio of free to total sulfur dioxide created a feature that is normally distributed.
```{r echo=FALSE, freetototalSO2_ratio}
#Calculating the ratio of free to total SO2.
wines$ftSO2ratio <- wines$free.sulfur.dioxide / wines$total.sulfur.dioxide

# This new feature has normal distribution.
ftSO2ratiop <- ggplot(aes(x = ftSO2ratio), data = wines) +
  geom_histogram(color = I('black'), fill = I('#7cc1b6')) +
  ggtitle('Free SO2 to Total SO2 Ratio') +
  theme(text = element_text(size=10))

grid.arrange(freeSO2p1, totalSO2p1, ftSO2ratiop, ncol = 3)
```

```{r echo=FALSE, citricacid_to_fixedacidity_ratio}
wines$cafaratio <- wines$citric.acid / wines$fixed.acidity

cafaratiop <- ggplot(aes(x = cafaratio), data = wines) +
  geom_histogram(color=I('black'), fill = I('#5d2cb2')) +
  theme(text = element_text(size = 10)) +
  ggtitle('Ratio of citric acid to fixed acidity')

# citric acid to fixed acidity ratio has the same distribution as the individual features.
```

### Boxplots of individual features

These plots show the spread of the data in another way, and show that all features have outliers.

```{r echo=FALSE, boxplots_univariate}
par(mfrow=c(1,6), oma = c(1,1,0,0) + 0.1,  mar = c(3,3,1,1) + 0.1)

boxplot(wines$fixed.acidity, col = "slategray2", pch=19)
mtext('Fixed Acidity', cex = 0.8, side = 1, line = 2)

boxplot(wines$volatile.acidity, col = "slategray2", pch=19)
mtext('Volatile Acidity', cex = 0.8, side = 1, line = 2)

boxplot(wines$citric.acid, col = "slategray2", pch=19)
mtext('Citric Acid', cex = 0.8, side = 1, line = 2)

boxplot(wines$residual.sugar, col = "slategray2", pch=19)
mtext('Residual Sugar', cex = 0.8, side = 1, line = 2)

boxplot(wines$chlorides, col = "slategray2", pch=19)
mtext('Chlorides', cex = 0.8, side = 1, line = 2)

boxplot(wines$free.sulfur.dioxide, col = "slategray2", pch=19)
mtext('Free Sulfur Dioxide', cex = 0.8, side = 1, line = 2)

boxplot(wines$total.sulfur.dioxide, col = "slategray2", pch=19)
mtext('Total Sulfur Dioxide', cex = 0.8, side = 1, line = 2)

boxplot(wines$density, col = "slategray2", pch=19)
mtext('Density', cex = 0.8, side = 1, line = 2)

boxplot(wines$pH, col = "slategray2", pch=19)
mtext('pH', cex = 0.8, side = 1, line = 2)

boxplot(wines$sulphates, col = "slategray2", pch=19)
mtext('Sulphates', cex = 0.8, side = 1, line = 2)

boxplot(wines$alcohol, col = "slategray2", pch=19)
mtext('Alcohol', cex = 0.8, side = 1, line = 2)
```

### Distribution of features by wine type
```{r echo=FALSE, histograms_by_type}
fixedacidityp3 <- ggplot(aes(x = fixed.acidity), data = wines) +
  geom_histogram(fill = I('#ddb40d'), color = I('black'), binwidth = 0.5) +
  scale_x_continuous(breaks = seq(4, 18, 2)) +
  ggtitle('Fixed Acidity in White (0) and Red (1) Wines') +
  facet_wrap( ~ type) +
  theme(text = element_text(size=10))
fixedacidityp3

volatileacidityp3 <- ggplot(aes(x = volatile.acidity), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#d9e570'), binwidth = 0.05) +
  scale_x_continuous(limits = c(0, 1)) +
  ggtitle('Volatile Acidity') +
  facet_wrap( ~ type) +
  theme(text = element_text(size=10))
volatileacidityp3

citricacidp3 <- ggplot(aes(x = citric.acid), data = wines) +
  geom_histogram(color = I('black'), fill = I ('orange')) +
  scale_x_continuous(limits = c(0, 1)) +
  ggtitle('Citric Acid') +
  facet_wrap( ~ type) +
  theme(text = element_text(size=10))
citricacidp3

residualsugarp3 <- ggplot(aes(x = residual.sugar), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#35a1ba'), binwidth = 0.2) +
  scale_x_continuous(breaks = seq(0, 7, 1), limits = c(0, 7)) +
  ggtitle('Residual Sugar') +
  facet_wrap( ~ type) +
  theme(text = element_text(size=10))
residualsugarp3

chloridesp3 <- ggplot(aes(x = chlorides), data = wines) +
  geom_histogram(color = I('black'), fill = I ('yellow'), binwidth = 0.005) +
  scale_x_continuous(limits = c(0.025, 0.15), breaks = seq(0.03, 0.15, 0.03)) +
  ggtitle('Chlorides') +
  facet_wrap( ~ type) +
  theme(text = element_text(size=10))
chloridesp3

freeSO2p3 <- ggplot(aes(x = free.sulfur.dioxide), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#35ba5d')) +
  ggtitle('Free Sulfur Dioxide') +
  scale_x_continuous(breaks = seq(0, 150, 50), limits = c(0, 150)) +
  facet_wrap( ~ type) +
  theme(text = element_text(size=10))
freeSO2p3

totalSO2p3 <- ggplot(aes(x = total.sulfur.dioxide), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#5d6d3a'), binwidth = 10) +
  scale_x_continuous(limits = c(0, 300)) +
  ggtitle('Total Sulfur Dioxide') +
  facet_wrap( ~ type) +
  theme(text = element_text(size=10))
totalSO2p3

densityp3 <- ggplot(aes(x = density), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#1f61c4'), binwidth = 0.001) +
  scale_x_continuous(limits = c(0.985, 1.005)) +
  facet_wrap( ~ type) +
  ggtitle('Density') +
  theme(text = element_text(size=10))
densityp3 

pHp3 <- ggplot(aes(x = pH), data = wines) +
  geom_histogram(color = I('black'), fill = I ('gray'), binwidth = 0.05) +
  facet_wrap( ~ type) +
  ggtitle('pH') +
  theme(text = element_text(size=10))
pHp3

SO4p3 <- ggplot(aes(x = sulphates), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#1fc42f'), binwidth = 0.05) +
  scale_x_continuous(limits = c(0.25,1.5)) +
  ggtitle('Sulphates') +
  facet_wrap( ~ type) +
  theme(text = element_text(size=10))
SO4p3

alcoholp3 <- ggplot(aes(x = alcohol), data = wines) +
  geom_histogram(color = I('black'), fill = I ('#727bd8'), binwidth = 0.2) +
  facet_wrap( ~ type) +
  ggtitle('Alcohol') +
  theme(text = element_text(size=10))
alcoholp3
```

### Boxplots of features according to wine type (Reds (1) and Whites(0))

```{r echo=FALSE}
p1 <- ggplot(aes(x = quality, y = fixed.acidity), data = wines) +
  geom_boxplot() +
  coord_cartesian(ylim = c(5, 12)) +
  facet_wrap(~ type) +
  ggtitle('Fixed Acidity') +
  theme(text = element_text(size=10))
p2 <- ggplot(aes(x = quality, y = volatile.acidity), data = wines) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0.1, 1.2)) +
  facet_wrap(~ type) +
  ggtitle('Volatile Acidity') +
  theme(text = element_text(size=10))
p3 <- ggplot(aes(x = quality, y = citric.acid), data = wines) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 0.6)) +
  facet_wrap(~ type) +
  ggtitle('Citric Acid') +
  theme(text = element_text(size=10))
p4 <- ggplot(aes(x = quality, y = residual.sugar), data = wines) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 12.5)) +
  facet_wrap(~ type) +
  ggtitle('Residual Sugar') +
  theme(text = element_text(size=10))
p5 <- ggplot(aes(x = quality, y = chlorides), data = wines) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 0.2)) +
  facet_wrap(~ type) +
  ggtitle('Chlorides') +
  theme(text = element_text(size=10))
p6 <- ggplot(aes(x = quality, y = free.sulfur.dioxide), data = wines) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 60)) +
  facet_wrap(~ type) +
  ggtitle('Free Sulfur Dioxide') +
  theme(text = element_text(size=10))
p7 <- ggplot(aes(x = quality, y = total.sulfur.dioxide), data = wines) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 225)) +
  facet_wrap(~ type) +
  ggtitle('Total Sulfur Dioxide') +
  theme(text = element_text(size=10))
p8 <- ggplot(aes(x = quality, y = density), data = wines) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0.985, 1)) +
  facet_wrap(~ type) +
  ggtitle('Density') +
  theme(text = element_text(size=10))
p9 <- ggplot(aes(x = quality, y = pH), data = wines) +
  geom_boxplot() +
  coord_cartesian(ylim = c(2.5, 3.75)) +
  facet_wrap(~ type) +
  ggtitle('pH') +
  theme(text = element_text(size=10))
p10 <- ggplot(aes(x = quality, y = sulphates), data = wines) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0.1, 0.9)) +
  facet_wrap(~ type) +
  ggtitle('Sulphates') +
  theme(text = element_text(size=10))
p11 <- ggplot(aes(x = quality, y = alcohol), data = wines) +
  geom_boxplot() +
  coord_cartesian(ylim = c(8, 14)) +
  facet_wrap(~ type) +
  ggtitle('Alcohol') +
  theme(text = element_text(size=10))

grid.arrange(p1, p2, p3, p4, ncol = 2)

grid.arrange(p5, p6, p7, p8, ncol = 2)

grid.arrange(p9, p10, p11, ncol = 2)

```


# Univariate Analysis

### What is the structure of your dataset?

The resulting "wines"" dataset has 6497 observations and 15 variables after creating it from the red wine and the white wine data sets downloaded from the Udacity project site. 

### What is/are the main feature(s) of interest in your dataset?

Main features are the quality and type of wines. Quality is dependent on a human test panel. It can be seen from the histograms that most wines were classified as 5, 6 and 7. Only a few made the 8 and the 9 classification (only five 9's and none of them were reds). There were 30 worst wines classified as 3. None of the wines were classified as 1, 2 and 10. 

From the boxplots, all features of wines have outliers. Removing some of the outliers in the plot, we can see a better distribution of the features in all of the wines and that red wines usually have a different distribution than white wines.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

Looking at the features separately in the red and white wines using facet_wrap function allowed us to see the difference in the distribution of the features in each type of wine. The distribution is different for the red and white wines in almost all features except pH and sulphates. It might be said that all the features can help classify white wines from red wines. 

Initial exploration of the features with respect to quality and wine type showed that some features have trends. An example is citric acid which seem to increase the quality of red wines, but does not do anything for white wines. Sulphates also seem to increase the quality of red wines, but not really for white wines.

However, it looks like a balance of some of the features can determine the quality of wines.

### Did you create any new variables from existing variables in the dataset?

Some of the features are correlated. From the information given, total sulfur dioxide is a measure of all sulfur dioxide, dissolved and undissolved. So it made sense to calculate the ratio of free sulfur dioxide to total sulfur dioxide and see what the distribution of this new feature was. Its histogram gave a normal distribution (see plots under transformation) while the individual features were not normally distributed. 

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

See plots under transformation.

All the non-normal distributions were transformed in some way. But some didn't change after transformation.

Fixed acidity is tailing, so transformation was done. The resulting histogram is more normally distributed.

Volatile acidity is skewed to the left and log10 transformation showed the bimodal characteristic of the distribution.

Residual sugar is not normally distributed. Transformation using log10 yielded something like a bimodal distribution.

Chlorides also don't look normally distributed. Transformation made it look better, but it also revealed a bimodal distribution.

Free sulfur dioxide and total sulfur dioxide had non-normal distribution and transformation didn't do anything. But when the ratio of the two have a normal distribution.

Transformation of density also didn't make the distribution better.

Alcohol distribution isn't normal and transformation didn't change the distribution.

Citric acid to fixed acidity ratio has the same distribution as the individual features.

# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
#ggpairs(wines, 
#  lower = list(continuous = wrap("points", shape = I('.'))), 
#  upper = list(combo = wrap("box", outlier.shape = I('.'))))
```

```{r echo=FALSE}
# Separating red wines from white wines
redwines <- subset(wines, type == 1)
whitewines <- subset(wines, type == 0)
```

### Quality vs. citric acid

```{r echo=FALSE}
# Converting back quality to a numeric class
wines$quality.num <- as.numeric(as.character(wines$quality))
```


```{r echo=FALSE}
ggplot(aes(x = quality.num, y = citric.acid), data = subset(wines, type == 1)) +
  geom_point(position = 'jitter', alpha=0.25, color = I('blue')) +
  geom_line(stat = 'summary', fun.y = median) +
  scale_x_continuous(breaks = c(3,4,5,6,7,8)) +
  ggtitle('Quality vs. citric acid in red wines') +
  theme(text = element_text(size = 10))

redwines <- subset(wines, type == 1)
cor.test(redwines$quality.num, redwines$citric.acid)
```

### Quality vs. chlorides

```{r echo=FALSE}
ggplot(aes(x = quality.num, y = chlorides), data = subset(wines, type == 1)) +
  geom_point(position = 'jitter', alpha=0.25, color = I('yellow')) +
  geom_line(stat = 'summary', fun.y = median) +
  scale_x_continuous(breaks = c(3,4,5,6,7,8)) +
  scale_y_continuous(limits = c(0, 0.2)) +
  ggtitle('Quality vs. chlorides in red wines') +
  theme(text = element_text(size = 10))

cor.test(redwines$quality.num, redwines$chlorides)
```

### Quality vs. volatile acidity

```{r echo=FALSE}
ggplot(aes(x = quality.num, y = volatile.acidity), data = subset(wines, type == 1)) +
  geom_point(position = 'jitter', alpha=0.25, color = I('orange')) +
  geom_line(stat = 'summary', fun.y = median) +
  scale_x_continuous(breaks = c(3,4,5,6,7,8)) +
  ggtitle('Quality vs. volatile acidity in red wines') +
  theme(text = element_text(size = 10))

cor.test(redwines$quality.num, redwines$volatile.acidity)
```

### Quality vs. pH

```{r echo=FALSE}
ggplot(aes(x = quality.num, y = pH), data = subset(wines, type == 1)) +
  geom_point(position = 'jitter', alpha=0.25, color = I('red')) +
  geom_line(stat = 'summary', fun.y = median) +
  scale_x_continuous(breaks = c(3,4,5,6,7,8)) +
  ggtitle('Quality vs. pH in red wines') +
  theme(text = element_text(size = 10))

cor.test(redwines$quality.num, redwines$pH)
```

### Quality vs. sulphates

```{r echo=FALSE}
ggplot(aes(x = quality.num, y = sulphates), data = subset(wines, type == 1)) +
  geom_point(position = 'jitter', alpha=0.25, color = I('#3a9960')) +
  geom_line(stat = 'summary', fun.y = median) +
  scale_x_continuous(breaks = c(3,4,5,6,7,8)) +
  ggtitle('Quality vs. sulphates in red wines') +
  theme(text = element_text(size = 10))

cor.test(redwines$quality.num, redwines$sulphates)
```


```{r echo=FALSE}
ggplot(aes(x = quality.num, y = ftSO2ratio), data = subset(wines, type == 1)) +
  geom_point(position = 'jitter', alpha=0.25, color = I('#3a9960')) +
  geom_line(stat = 'summary', fun.y = mean) +
  scale_x_continuous(breaks = c(3,4,5,6,7,8)) +
  ggtitle('Quality vs. free/total SO2 in red wines') +
  theme(text = element_text(size = 10))
```


### Quality vs. alchohol

```{r echo=FALSE}
ggplot(aes(x = quality.num, y = alcohol), data = subset(wines, type == 1)) +
  geom_point(position = 'jitter', alpha=0.25, color = I('#4b67bc')) +
  geom_line(stat = 'summary', fun.y = median) +
  scale_x_continuous(breaks = c(3,4,5,6,7,8)) +
  ggtitle('Quality vs. alcohol in red wines') +
  theme(text = element_text(size = 10))

cor.test(redwines$quality.num, redwines$alcohol)
```

### Quality vs. density

```{r echo=FALSE}
ggplot(aes(x = quality.num, y = density), data = subset(wines, type == 1)) +
  geom_point(position = 'jitter', alpha=0.25, color = I('#b52f81')) +
  geom_line(stat = 'summary', fun.y = median) +
  scale_x_continuous(breaks = c(3,4,5,6,7,8)) +
  ggtitle('Quality vs. density in red wines') +
  theme(text = element_text(size = 10))

cor.test(redwines$quality.num, redwines$density)
```


### Citric acid vs. pH

```{r echo=FALSE, citric.acid_pH}
ggplot(aes(x = citric.acid, y = pH), data = redwines) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = 'black') +
  facet_wrap( ~ quality) +
  theme(text = element_text(size = 10)) +
  scale_color_brewer(type = 'qual') +
  ggtitle('Citric acid vs. pH in Red Wines')

cor.test(redwines$pH, redwines$citric.acid)

ggplot(aes(x = citric.acid, y = pH), data = whitewines) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = 'black') +
  facet_wrap( ~ quality) +
  theme(text = element_text(size = 10)) +
  scale_color_brewer(type = 'qual') +
  ggtitle('Citric acid vs. pH in White Wines')

cor.test(whitewines$pH, whitewines$citric.acid)
```


### Volatile acidity vs. pH

```{r echo=FALSE, volatileacidity_pH}
ggplot(aes(x = volatile.acidity, y = pH), data = redwines) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = 'black') +
  facet_wrap( ~ quality) +
  scale_color_brewer(type = 'qual') +
  ggtitle('Volatile acidity vs. pH in Red Wines') +
  theme(text = element_text(size = 10)) 

cor.test(redwines$volatile.acidity, redwines$pH)

ggplot(aes(x = volatile.acidity, y = pH), data = whitewines) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = 'black') +
  facet_wrap( ~ quality) +
  scale_color_brewer(type = 'qual') +
  ggtitle('Volatile acidity vs. pH in White Wines') +
  theme(text = element_text(size = 10)) 

cor.test(whitewines$volatile.acidity, whitewines$pH)
```

### Citric acid vs. volatile acidity

```{r echo=FALSE}
ggplot(aes(x = citric.acid, y = volatile.acidity), data = redwines) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = '#595b5b') +
  facet_wrap( ~ quality) +
  scale_color_brewer(type = 'qual') +
  ggtitle('Citric acid vs. volatile acidity in Red Wines') +
  theme(text = element_text(size = 10)) 

cor.test(redwines$citric.acid, redwines$volatile.acidity)

ggplot(aes(x = citric.acid, y = volatile.acidity), data = whitewines) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = '#595b5b') +
  facet_wrap( ~ quality) +
  scale_color_brewer(type = 'qual') +
  ggtitle('Citric acid vs. volatile acidity in white Wines') +
  theme(text = element_text(size = 10)) 

cor.test(whitewines$citric.acid, whitewines$volatile.acidity)
```

### Density vs alcohol

```{r echo=FALSE, density_alcohol}
ggplot(aes(x = density, y = alcohol), data = redwines) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = '#595b5b') +
  facet_wrap( ~ quality) +
  scale_color_brewer(type = 'qual') +
  ggtitle('Density vs. Alcohol in Red Wines') +
  theme(text = element_text(size = 10))

cor.test(redwines$density, redwines$alcohol)
  
ggplot(aes(x = density, y = alcohol), data = subset(whitewines, density <= 1)) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = '#595b5b') +
  facet_wrap( ~ quality) +
  scale_color_brewer(type = 'qual') +
  ggtitle('Density vs. Alcohol in White Wines') +
  theme(text = element_text(size = 10))

cor.test(whitewines$density, whitewines$alcohol)
```

### Density vs. residual sugar

```{r echo=FALSE, residual.sugar_density}
ggplot(aes(x = density, y = residual.sugar), data = subset(redwines, density = quantile(density, 0.99))) +
  geom_point(aes(color=quality)) +
  geom_smooth(method = 'lm', color = '#595b5b') +
  scale_y_continuous(limits = c(0, 8)) +
  facet_wrap(~quality, ncol = 2) +
  ggtitle('Density vs. residual sugar in red wines') +
  theme(text = element_text(size = 10)) +
  scale_color_brewer(type = 'qual')

cor.test(redwines$density, redwines$residual.sugar)

ggplot(aes(x = density, y = residual.sugar), data = subset(whitewines, density = quantile(density, 0.99))) +
  geom_point(aes(color=quality)) +
  geom_smooth(method = 'lm', color = '#595b5b') +
  scale_y_continuous(limits = c(0, 8)) +
  scale_x_continuous(limits = c(0.985, 1.0)) +
  facet_wrap(~quality, ncol = 2) +
  ggtitle('Density vs. residual sugar in white wines') +
  theme(text = element_text(size = 10)) +
  scale_color_brewer(type = 'qual')

cor.test(whitewines$density, whitewines$residual.sugar)
```

Effect of pH on free sulfur dioxide

```{r echo=FALSE}
ggplot(aes(x = pH, y = ftSO2ratio), data = redwines) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = '#595b5b') +
  facet_wrap( ~ quality) +
  theme(text = element_text(size = 10)) +
  scale_color_brewer(type = 'qual')

cor.test(redwines$pH, redwines$ftSO2ratio)

ggplot(aes(x = pH, y = ftSO2ratio), data = whitewines) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = '#595b5b') +
  facet_wrap( ~ quality) +
  theme(text = element_text(size = 10)) +
  scale_color_brewer(type = 'qual')

cor.test(whitewines$pH, whitewines$ftSO2ratio)
```

Effect of pH on chlorides

```{r echo=FALSE}
ggplot(aes(x = pH, y = chlorides), data = subset(redwines, chlorides = quantle(chlorides, 0.95))) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = '#595b5b') +
  facet_wrap( ~ quality) +
  theme(text = element_text(size = 10)) +
  scale_color_brewer(type = 'qual')

cor.test(redwines$pH, redwines$chlorides)

ggplot(aes(x = pH, y = chlorides), data = subset(whitewines, chlorides = quantle(chlorides, 0.95))) +
  geom_point(aes(color = quality)) +
  geom_smooth(method = 'lm', color = '#595b5b') +
  facet_wrap( ~ quality) +
  theme(text = element_text(size = 10)) +
  scale_color_brewer(type = 'qual')

cor.test(whitewines$pH, whitewines$chlorides, method = "spearman")
```

### Correlation matrix
```{r echo=FALSE}
#cor(whitewines[,-12])
#cor(whitewines[,-12], method="spearman")
pairs(wines[,-12], gap=0, pch=19, cex=0.4, col="darkblue")
title(sub="Scatterplot of Chemical Attributes", cex=0.8)
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

Trends were seen between quality and a few features for red wines but not for white wines. The following are observations from the red wines:

a. Volatile acidity seems to decrease the quality of red wines.
b. Citric acid seems to increase red wine quality. This is supported by the trend seen in pH of red wines, though not too much by fixed acidity. If there is more citric acid, the pH would expectedly be lower. So in the boxplot for pH, the best quality wine have the lowest mean pH.
c. Quality increases with sulphates.
d. Quality also increases with alcohol.
e. Quality decreases with density, which makes sense. When alcohol is higher, the less dense the wine is expected to be.
f. The best wines have the lowest chlorides.
g. Sulfur dioxide (free and total) does not seem to affect quality.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

As mentioned in the information provided with the data set, some features are correlated. For example, pH is definitely correlated to citric acid. The more citric acid there is, pH would expectedly be lower. The scatterplots and Pearson correlation coefficients of pH vs. citric acid, pH vs. volatile acidity and citric acid vs. volatile acidity showed these relationships, (if any) in both the reds and the whites.

There is a correlation between residual sugar and density, which also makes sense, since the more sugar there are in the wine, the more dense it is expected to be. This is specially seen in the scatterplots for white wines.

Density and alcohol are also expected to be related since alcohol is less dense than water, the more alcohol there is in wine, the less dense wine is expected to be. 

Other relationships were examined. The effect of pH on chlorides were examined by looking at their scatterplots. Though using all wines, there is no correlation between the two features, looking at only a few, for example, red wines with quality 8, a good (negative) correlation between the two can be found (the higher the pH, the lower the chlorides which also makes sense since a more basic pH means less ioned salt.)

### What was the strongest relationship you found?

```{r echo=FALSE}
redscors <- cor(redwine)
redscors
whitescors <- cor(whitewine)
whitescors
```

For all wines, between quality and any of the other features, alcohol had the highest Pearson correlation coefficient.
```{r echo=FALSE}
#redscors >= 0.3
#whitescors >= 0.3
```

The highest correlations are found between:

- residual.sugar and density in white wines
- citric.acid and fixed.acidity in red wines
- total and free sulfur dioxide in both wines

# Multivariate Plots Section

### Histograms
```{r echo=FALSE}
fixedacidityp4 <- ggplot(aes(x = fixed.acidity), data = wines) +
  geom_histogram(aes(fill = quality), binwidth = 0.5) +
  scale_x_continuous(breaks = seq(4, 18, 1)) +
  scale_fill_brewer(type='qual') +
  ggtitle('Fixed Acidity') +
  facet_wrap(~ type) +
  theme(text = element_text(size=10)) 

volatileacidityp4 <- ggplot(aes(x = volatile.acidity), data = wines) +
  geom_histogram(aes(fill = quality)) +
  scale_fill_brewer(type='qual') +
  ggtitle('Volatile Acidity') +
  facet_wrap(~ type) +
  theme(text = element_text(size=10))

citricacidp4 <- ggplot(aes(x = citric.acid), data = wines) +
  geom_histogram(aes(fill = quality), binwidth = 0.05) +
  scale_x_continuous(breaks = seq(0, 1, 0.2)) +
  scale_fill_brewer(type='qual') +
  ggtitle('Citric Acid') +
  facet_wrap(~ type) +
  theme(text = element_text(size=10))

residualsugarp4 <- ggplot(aes(x = residual.sugar), data = wines) +
  geom_histogram(aes(fill = quality), binwidth = 0.2) +
  scale_x_continuous(breaks = seq(0, 7, 1), limits = c(0, 7)) +
  ggtitle('Residual Sugar in Red Wines') +
  scale_fill_brewer(type='qual') +
  facet_wrap(~ type) +
  theme(text = element_text(size=10))

chloridesp4 <- ggplot(aes(x = chlorides), data = wines) +
  geom_histogram(aes(fill = quality), binwidth = 0.005) +
  scale_x_continuous(limits = c(0.025, 0.15), breaks = seq(0.03, 0.15, 0.03)) +
  scale_fill_brewer(type='qual') +
  facet_wrap(~ type) +
  ggtitle('Chlorides') +
  theme(text = element_text(size=10))

freeSO2p4 <- ggplot(aes(x = free.sulfur.dioxide), data = wines) +
  geom_histogram(aes(fill = quality)) +
  scale_fill_brewer(type='qual') +
  scale_x_continuous(breaks = seq(0, 70, 35)) +
  facet_wrap(~ type) +
  ggtitle('Free Sulfur Dioxide in Red Wines') +
  theme(text = element_text(size=10))

totalSO2p4 <- ggplot(aes(x = total.sulfur.dioxide), data = wines) +
  geom_histogram(aes(fill=quality), binwidth = 10) +
  scale_fill_brewer(type='qual') +
  scale_x_continuous(breaks = seq(0, 300, 50)) +
  ggtitle('Total Sulfur Dioxide') +
  facet_wrap(~ type) +
  theme(text = element_text(size=10))

densityp4 <- ggplot(aes(x = density), data = wines) +
  geom_histogram(aes(fill=quality), binwidth = 0.001) +
  scale_x_continuous(breaks = seq(0.98, 1.005, 0.01)) +
  scale_fill_brewer(type='qual') +
  facet_wrap(~ type) +
  ggtitle('Density') +
  theme(text = element_text(size=10))

pHp4 <- ggplot(aes(x = pH), data = wines) +
  geom_histogram(aes(fill = quality), binwidth = 0.05) +
  ggtitle('pH of Red Wines') +
  scale_fill_brewer(type='qual') +
  facet_wrap(~ type) +
  ggtitle('pH') +
  theme(text = element_text(size=10))

SO4p4 <- ggplot(aes(x = sulphates), data = wines) +
  geom_histogram(aes(fill = quality), binwidth = 0.05) +
  scale_fill_brewer(type='qual') +
  scale_x_continuous(breaks = seq(0, 2, 0.25), limits = c(0, 1.25)) +
  facet_wrap(~ type) +
  ggtitle('Sulphates') +
  theme(text = element_text(size=10))

alcoholp4 <- ggplot(aes(x = alcohol), data = wines) +
  geom_histogram(aes(fill = quality)) +
  scale_x_continuous(breaks = seq(8, 15, 1)) +
  scale_fill_brewer(type='qual') +
  facet_wrap(~ type) +
  ggtitle('Alcohol') +
  theme(text = element_text(size=10))

```

```{r echo=FALSE, All_Params}
grid.arrange(citricacidp4, SO4p4, pHp4)
```

### Scatterplots

Citric acid

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = citric.acid), data = subset(wines, citric.acid = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_y_continuous(limits = c(0, 0.8)) +
  scale_x_sqrt(breaks = seq(8, 14, 0.5)) +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))

# good, not too much
```

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = citric.acid), data = subset(wines, citric.acid = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_x_sqrt(breaks = seq(8, 14, 0.5)) +
  scale_y_log10() +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))

# not very good
```

Fixed acidity

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = fixed.acidity), data = subset(wines, fixed.acidity = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_y_continuous(limits = c(3, 12)) +
  scale_x_sqrt(breaks = seq(8, 14, 0.5)) +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))

# good
```

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = sqrt(fixed.acidity)), data = subset(wines, fixed.acidity = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_x_sqrt(breaks = seq(8, 14, 0.5)) +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))
# not good
```

Volatile acidity

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = volatile.acidity), data = subset(wines, volatile.acidity = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_y_continuous(limits = c(0.1, 1)) +
  scale_x_sqrt(breaks = seq(8, 14, 0.5)) +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))

# good
```

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = log10(volatile.acidity)), data = subset(wines, volatile.acidity = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_x_sqrt(breaks = seq(8, 14, 0.5)) +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))

# nope
```

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = density), data = subset(wines, density = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_x_sqrt(breaks = seq(8, 14, 0.5)) +
  scale_y_continuous(limits = c(0.985, 1.005)) +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))

# nope
```

Free SO2

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = free.sulfur.dioxide), data = subset(wines, free.sulfur.dioxide = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_x_sqrt(breaks = seq(8, 14, 0.5)) +
  scale_y_continuous(limits = c(0, 100)) + 
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))

```

Total SO2

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = total.sulfur.dioxide), data = subset(wines, total.sulfur.dioxide = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_x_sqrt(breaks = seq(8, 14, 0.5)) +
    scale_y_continuous(limits = c(0, 260)) + 
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))

```

Free to total SO2 ratio

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = ftSO2ratio), data = subset(wines, ftSO2ratio = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_x_sqrt() +
  scale_y_sqrt() +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))

```

Residual Sugar

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = residual.sugar), data = subset(wines, residual.sugar = quartiles(wines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_x_sqrt() +
  scale_y_continuous(limits = c(0, 20)) +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type) +
  theme(text = element_text(size = 10))

# good
```

```{r}
ggplot(aes(x = alcohol, y = residual.sugar), data = subset(redwines, residual.sugar = quartile(redwines, 0.9))) +
  geom_point(aes(color = quality)) +
  scale_x_sqrt() +
  scale_y_continuous(limits = c(0, 5)) +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr')
```

Density

```{r}
ggplot(aes(x = alcohol, y = density), data = subset(wines, density = quartile(wines, 0.99))) +
  geom_point(aes(color = quality)) +
  scale_x_sqrt() +
  scale_y_sqrt(limits = c(0.985, 1.005)) + 
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type)
# not that good
```


```{r}
ggplot(aes(x = density, y = free.sulfur.dioxide), data = subset(wines, density = quartile(wines, 0.99))) +
  geom_point(aes(color = quality)) +
  scale_x_continuous(limits = c(0.985, 1.005)) +
  scale_y_continuous(limits = c(0, 100)) + 
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type)
# not good
```

```{r}
ggplot(aes(x = sulphates, y = alcohol), data = subset(wines, alcohol = quartile(wines, 0.99))) +
  geom_point(aes(color = quality)) +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type)
# not good
```

```{r}
ggplot(aes(x = log10(chlorides+1), y = sqrt(alcohol)), data = subset(wines, alcohol = quartile(wines, 0.99))) +
  geom_point(aes(color = quality)) +
  scale_x_continuous(limits = c(0, 0.05)) +
  scale_y_continuous(limits = c(2.9, 3.8)) +
  scale_color_brewer(type = 'seq', palette = 'YlOrBr') +
  facet_wrap(~ type)
# not good
```
# Multivariate Analysis

#### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

There were some features that helped with alcohol content in classifying wines more than others. For example, volatile acidity helped more than density. This is because density and alcohol correlate with each other so the effects of the two features would be the same, causing the least separation among wine qualities. However, I wasn't able to find a feature that totally helped to separate the qualities of wines. 

#### Were there any interesting or surprising interactions between features?

The only thing that surprised me is that the ratio of free SO2 to total SO2 is normally distributed while the individual features are not. Total SO2 actually contributes more to the separation of the qualities of wine in the multivariate analysis. 

#### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

#### Plot One
```{r echo=FALSE, Plot_One}

```

#### Description One


#### Plot Two
```{r echo=FALSE, Plot_Two}

```

#### Description Two


#### Plot Three
```{r echo=FALSE, Plot_Three}

```

#### Description Three

------

# Reflection


# References

https://onlinecourses.science.psu.edu/stat857/print/book/export/html/224
http://www.practicalwinery.com/janfeb09/page2.htm


P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

----------
